---
title: "Introduction to ggplot: course code"
author: "Dr. Laurie Baker"
date: "28/04/2020"
output: 
  html_document:
    number_sections: true
    toc: yes
    toc_depth: '2'
    toc_float: yes
---

# Getting started

## Installing and loading packages

```{r libraries, echo=TRUE, message=TRUE, warning=TRUE}
# install.packages("tidyverse")
# install.packages("gapminder")

library(tidyverse)
library(gapminder)

```

## Cleaning and prepping the data

### This is messy data

```{r tidy-messy1, results='asis', echo=FALSE}

pop_simple <- gapminder %>% 
  filter(
    country %in% c("Syria", "Rwanda", "Chile"), 
    year >= 1997
  ) %>% 
  select(country, year, pop) %>% 
  mutate(pop = pop / 10^6)

messy_pop <- pop_simple %>% spread(year, pop)

knitr::kable(messy_pop, format = 'html')

```

### This is clean data

```{r tidy-tidy1, result='asis'}

tidy_pop <- gather(messy_pop, 'year', 'pop', -country)
tidy_pop %>%
  head(4) 

```


## Building a plot

### Calling `ggplot()`

```{r first-plot1a, ref.label='first-plot1a', out.width="100%"}

ggplot(tidy_pop)

```

### Adding aesthetics

```{r first-plot1b, ref.label='first-plot1b', out.width="100%"}

ggplot(tidy_pop,
  aes(x = year, #<<
      y = pop) #<<
  )
```

### Choosing the geometry (geom)

```{r first-plot1c, ref.label='first-plot1c', out.width="100%"}

ggplot(tidy_pop,
  aes(x = year,
      y = pop)
  ) +
  geom_point() 

```

### Adding a third variable, country, as a colour.

```{r first-plot1d, out.width = "100%"}

ggplot(tidy_pop,
  aes(x = year,
      y = pop,
      color = country) 
  ) + 
  geom_point()

```

### Adding a line to the plot

```{r first-plot1e, out.width = "100%"}

ggplot(tidy_pop,
  aes(x = year,
      y = pop,
      color = country) 
  ) +
  geom_point() +
  geom_line()

```

### Adding a line to the plot, group by country.

```{r first-plot1f, out.width = "100%"}

ggplot(tidy_pop,
  aes(x = year,
      y = pop,
      color = country) 
  ) +
  geom_point() +
  geom_line(
    aes(group = country)) 

```

### Storing a plot

```{r first-plot1g, out.width = "100%"}

g <- ggplot(tidy_pop, 
  aes(x = year,
      y = pop,
      color = country) 
  ) +
  geom_point() +
  geom_line(
    aes(group = country))

g 


```

### Specifying the aesthetics for each geometry.

```{r}

g <- ggplot() +
  geom_point(
    data = tidy_pop, 
    aes(x = year, 
      y = pop, 
      color = country) 
  ) +
  geom_line(
    data = tidy_pop, 
    aes(x = year, 
      y = pop, 
      color = country, 
      group = country) 
    )

g

```

### Adding more data. Creating `tidy_pop2`

```{r}

tidy_pop2 <- left_join(tidy_pop, select(gapminder, country, continent))

g <- ggplot(tidy_pop2) +
  aes(x = year,
      y = pop,
      color = country) +
  geom_point() +
  geom_line(aes(group = country))

```

### Using Facets

```{r geom_facet}

g + facet_wrap(~ country)

```

### Using facet_grid, two variables.

```{r geom_grid}
g + facet_grid(continent ~ country)
```

### Adding labels

```{r labs-ex}

(g <- g + labs(x = "Year", y = "Population (millions)"))

```

### Showing coordinates

```{r adding coordinates}

g + coord_flip()

```

### Changing the color scale

```{r}

g + scale_color_manual(values = c("peru", "pink", "plum"))

```

### Changing the scale of the y

```{r}

g + scale_y_log10()

```

### Changing the scale of the x

```{r}

g + scale_x_discrete(labels = c("MCMXCVII", "MMII", "MMVII"))

```

### Changing the theme

```{r theme ex1}

g + theme_bw()

```

### Changing the theme (fonts and colours) 

```{r theme ex2}

g + theme_bw() + theme(text = element_text(colour = "hotpink", size = 20))

```

### Setting an overall theme:

```{r Setting a theme}

my_theme <- theme_bw() +
  theme(
    text = element_text(size = 12),
    panel.border = element_rect(colour = 'grey80'), 
    panel.grid.minor = element_blank()
  )

theme_set(my_theme)

```


### Making the plot with the new theme

```{r}

theme_set(my_theme)
g 

```

### Changing the position of the legend

```{r}

theme_set(my_theme)
g + theme(legend.position = 'bottom')

```


# Live coding, examples using the gapminder data set


# "Live" Coding

Resetting the plot theme

```{r reset_theme, include=FALSE}
theme_set(theme_gray())
```

## head(gapminder)

```{r head-gapminder, echo=FALSE}
head(gapminder)
```

## glimpse(gapminder)

```{r summary-gapminder, echo=FALSE, comment=""}
glimpse(gapminder)
```

## Let's start by plotting `lifeExp` vs `gdpPercap`


```{r gapminder-le-gdp-1}
ggplot(gapminder,
  aes(x = gdpPercap,
      y = lifeExp))
```


### Add points...

```{r gapminder-le-gdp-2}
ggplot(gapminder,
  aes(x = gdpPercap,
      y = lifeExp)) +
  geom_point() #<<
```


### How can I tell countries apart?

Color by continent

```{r gapminder-le-gdp-3}
ggplot(gapminder,
  aes(x = gdpPercap,
      y = lifeExp,
      color = continent)) +  #<<
  geom_point()
```

### GDP is squished together on the left, let's change the axes.


```{r gapminder-le-gdp-4}
ggplot(gapminder,
  aes(x = gdpPercap,
      y = lifeExp,
      color = continent)) +
  geom_point() +
  scale_x_log10() #<<
```


### Still lots of overlap in the countries...


```{r gapminder-le-gdp-5}
ggplot(gapminder,
  aes(x = gdpPercap,
      y = lifeExp,
      color = continent)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~ continent) + #<<
  guides(color = FALSE)     #<<
```

No need for color legend thanks to facet titles


### Lots of overplotting due to point size, let's make the points smaller


```{r gapminder-le-gdp-6}
ggplot(gapminder,
  aes(x = gdpPercap,
      y = lifeExp,
      color = continent)) +
  geom_point(size = 0.25) + #<<
  scale_x_log10() +
  facet_wrap(~ continent) +
  guides(color = FALSE)
```


Is there a trend?


```{r gapminder-le-gdp-7}
ggplot(gapminder,
  aes(x = gdpPercap,
      y = lifeExp,
      color = continent)) +
  geom_line() + #<<
  #geom_point(size = 0.25) +
  scale_x_log10() +
  facet_wrap(~ continent) +
  guides(color = FALSE)
```

Okay, that line just connected all of the points sequentially...

### Let's add the group as an aesthetic to `geom_line()`

```{r gapminder-le-gdp-8}
ggplot(gapminder,
  aes(x = gdpPercap,
      y = lifeExp,
      color = continent)) +
  geom_line(
    aes(group = country) #<<
    ) +
  #geom_point(size = 0.25) +
  scale_x_log10() +
  facet_wrap(~ continent) +
  guides(color = FALSE)
```

### Let's put year on the x-axis to look at the time series.


```{r gapminder-gdp-year-1}
ggplot(gapminder,
  aes(x = year, #<<
      y = gdpPercap, #<<
      color = continent)) +
  geom_line(
    aes(group = country)) +
  #geom_point(size = 0.25) +
  scale_y_log10() + #<<
  facet_wrap(~ continent) +
  guides(color = FALSE)
```


### Can't see x-axis labels, let's change the x axis


```{r gapminder-gdp-year-2}
ggplot(gapminder,
  aes(x = year,
      y = gdpPercap,
      color = continent)) +
  geom_line(
    aes(group = country)) +
  #geom_point(size = 0.25) +
  scale_y_log10() +
  scale_x_continuous(breaks = #<<
    seq(1950, 2000, 25) #<<
  ) +                            #<<
  facet_wrap(~ continent) +
  guides(color = FALSE)
```


### What about life expectancy?


```{r gapminder-le-year-1}
ggplot(gapminder, 
  aes(x = year, 
      y = lifeExp, #<<
      color = continent)) +
  geom_line(
    aes(group = country)) +
  #geom_point(size = 0.25) +
  #scale_y_log10() + #<<
  scale_x_continuous(breaks = 
    seq(1950, 2000, 25)) +  
  facet_wrap(~ continent) +
  guides(color = FALSE)
```


Okay, let's add a trend line

### Adding a trendline with geom_smooth()

```{r gapminder-le-year-2}
ggplot(gapminder,
  aes(x = year, 
      y = lifeExp,
      color = continent)) +
  geom_line(
    aes(group = country)) +
  # geom_point(size = 0.25) +
  geom_smooth() + #<<
  scale_x_continuous(breaks = 
    seq(1950, 2000, 25)) +  
  facet_wrap(~ continent) +
  guides(color = FALSE)
```


### De-emphasize individual countries using alpha


```{r gapminder-le-year-3}
ggplot(gapminder,
  aes(x = year, 
      y = lifeExp,
      color = continent)) +
  geom_line(
    aes(group = country),
    alpha = 0.2) + #<<
  #geom_point(size = 0.25) +
  geom_smooth() + 
  scale_x_continuous(breaks = 
    seq(1950, 2000, 25)) +  
  facet_wrap(~ continent) +
  guides(color = FALSE)
```


### Let's compare continents

```{r gapminder-le-year-5}
ggplot(gapminder,
  aes(x = year, 
      y = lifeExp,
      color = continent)) +
  geom_line(
    aes(group = country),
    alpha = 0.2) +
  geom_smooth() + 
  # scale_x_continuous(
  #   breaks = 
  #     seq(1950, 2000, 25))+
  # facet_wrap(~ continent) + #<<
  guides(color = FALSE)
```
]

Wait, what color is each continent?

### Adding a legend

```{r gapminder-le-year-6}
ggplot(gapminder, 
  aes(x = year, 
      y = lifeExp,
      color = continent)) +
  geom_line(
    aes(group = country),
    alpha = 0.2) +
  geom_smooth() + 
  theme( #<<
  legend.position = "bottom") #<<
```


### Let's try adding the minimal theme


```{r gapminder-le-year-7}
ggplot(gapminder, 
  aes(x = year, 
      y = lifeExp,
      color = continent)) +
  geom_line(
    aes(group = country),
    alpha = 0.2) +
  geom_smooth() + 
  theme_minimal() + #<<
  theme(
  legend.position = "bottom")
```

Fonts are kind of big

### Can we make the font smaller?

```{r gapminder-le-year-8}
ggplot(gapminder, 
  aes(x = year, 
      y = lifeExp,
      color = continent)) +
  geom_line(
    aes(group = country),
    alpha = 0.2) +
  geom_smooth() + 
  theme_minimal( 
    base_size = 10) + #<<
  theme(
  legend.position = "bottom")
```

Great, let's switch gears

### Let's explore a few countries in more detail

```{r gapminder-americas-data}
americas <- 
  gapminder %>% 
  filter(
    country %in% c(
      "Brazil",
      "Canada",
      "Mexico",
      "Ecuador"
    )
  )
```

### Let's look at four countries in more detail. How do their populations compare to each other?


```{r echo=FALSE}
americas %>%
  head(10)
```

### Making a bar chart

```{r gapminder-americas-1}
ggplot(americas,
  aes(x = year,
      y = pop) 
  ) +
  geom_col()
```



Yeah, but how many people are in each country?

### Color the bars by country

```{r gapminder-americas-2}
ggplot(americas,
  aes(x = year,
      y = pop,
      fill = country) #<<
  ) +
  geom_col()
```

Can we reorder by population size?

### Using fct_reorder() from the forcats package

Excellent tutorial by Jenny Bryan ["Be the Boss of your factors"](https://stat545.com/factors-boss.html)

```{r gapminder-americas-2-1}

ggplot(americas,
  aes(x = year,
      y = pop,
      fill = fct_reorder( #<< 
        country, pop) #<< 
      )
  ) +
  geom_col()
```

Can we change the labels?


### Add a label for country.

```{r gapminder-americas-2-2}

ggplot(americas,
  aes(x = year,
      y = pop,
      fill = fct_reorder( #<< 
        country, pop) #<< 
      )
  ) +
  geom_col() +
  labs(fill = "Country") #<< 
```

Bars are "stacked", can we separate?

### Changing the position using "dodge"

```{r gapminder-americas-3}
ggplot(americas,
  aes(x = year,
      y = pop,
      fill = fct_reorder(  
        country, pop) 
      )
  ) +
  geom_col(
    position = "dodge") + #<<
  labs(fill = "Country") 
```

`position = "dodge"` places objects *next to each other* instead of overlapping

What is scientific notation anyway?


### Dividing population to be smaller

```{r gapminder-americas-4}
ggplot(americas,
  aes(x = year,
      y = pop / 10^6, #<<
      fill = fct_reorder(  
        country, pop)
      ) 
  ) +
  geom_col(
    position = "dodge" 
  ) +
  labs(fill = "Country", 
       y = "Population (millions)"#<<
       )  
```

ggplot aesthetics can take expressions! Still, it might be easier to see countries individually



### Using facet_wrap() to create individual plots


```{r gapminder-americas-5}
ggplot(americas,
  aes(x = year,
      y = pop / 10^6, #<<
      fill = country) #<<
  ) +
  geom_col(
    position = "dodge" 
  ) +
  labs(y = "Population (millions)") + 
  facet_wrap(~ country) + #<<
  guides(fill = FALSE) #<<
```

Let range of y-axis vary in each plot


### Changing the scales

```{r gapminder-americas-6}
ggplot(americas,
  aes(x = year,
      y = pop / 10^6, 
      fill = country)
  ) +
  geom_col(
    position = "dodge" 
  ) +
  labs(y = "Population (millions)") + 
  facet_wrap(~ country,
    scales = "free_y") + #<<
  guides(fill = FALSE) 
```

What about life expectancy again?


### Making a barchart with life expectancy

```{r gapminder-americas-7}
ggplot(americas,
  aes(x = year,
      y = lifeExp, #<<
      fill = country)
  ) +
  geom_col(
    position = "dodge" 
  ) +
  facet_wrap(~ country,
             scales = "free_y") +
  guides(fill = FALSE) 
```

This should really be a line instead of a bar

### Changing the plot to a line plot

```{r gapminder-americas-8}
ggplot(americas,
  aes(x = year,
      y = lifeExp, 
      fill = country)
  ) +
  geom_line() + #<<
  facet_wrap(~ country,
             scales = "free_y") +
  guides(fill = FALSE)
```


### Coloring each line by country. No need to show in the legend as each country has it's own plot.

```{r gapminder-americas-9}
ggplot(americas) +
  aes(
    x = year,
    y = lifeExp,
    color = country #<<
  ) +
  geom_line() +
  facet_wrap(~ country,
    scales = "free_y") +
  guides(color = FALSE) #<<
```

Altogether now!

### Putting all the plots back together again

```{r gapminder-americas-10}
ggplot(americas) +
  aes(
    x = year,
    y = lifeExp,
    color = country
  ) +
  geom_line()
```

Let's update the labels

### Adding labels to the plots

```{r gapminder-americas-11}
ggplot(americas,
  aes(x = year,
      y = lifeExp,
      color = country)
  ) +
  geom_line() +
  labs(x = "Year",
       y = "Life expectancy (years)",
       color = "Country",
       title = "Life expectancy over time")
```

Okay, changing gears again. What is range of life expectancy in Americas?


### Combining dplyr and ggplot code!

```{r gapminder-all-americas-1}
gapminder %>% 
  filter(
    continent == "Americas"
  ) %>% #<<
  ggplot( #<<
    aes(x = year,
        y = lifeExp) 
  )
```

You can pipe into `ggplot()`!  
Just watch for `%>%` changing to `+`


## Creating a box plot

Boxplot for life expectancy range

```{r gapminder-all-americas-2}
gapminder %>% 
  filter(
    continent == "Americas"
  ) %>% 
  ggplot( 
    aes(x = year, 
        y = lifeExp) 
  ) + 
  geom_boxplot() #<<
```


Why not boxplots by year?

### Creating the box plots by year

```{r gapminder-all-americas-3}
gapminder %>% 
  filter(
    continent == "Americas"
  ) %>%
  mutate( #<<
    year = factor(year) #<<
  ) %>%  #<<
  ggplot( 
    aes(x = year, 
        y = lifeExp) 
  ) + 
  geom_boxplot()
```


### Changing the coordinates using coord_flip()

```{r gapminder-all-americas-4}
gapminder %>% 
  filter(
    continent == "Americas"
  ) %>%
  mutate(
    year = factor(year) 
  ) %>%  
  ggplot( 
    aes(x = year, 
        y = lifeExp) 
  ) + 
  geom_boxplot() +
  coord_flip() #<<
```
